---
title: "ixsurface: Function Reference and Examples"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
install.packages(".", repos = NULL, type = "source", quiet = TRUE)
library(ixsurface)
`%>%` = plotly::`%>%`
```

# Overview

`ixsurface` generates interactive 3D surface plots from fitted models to visualize
multi-factor interactions. Where surfaces cross, the effect of one factor depends
on the level of another — the geometric signature of an interaction.

This document walks through every exported function, its parameters, and output.

---

# sim_factorial

Generates synthetic factorial experiment data with known interaction structure.
Useful for testing and demonstration.

## Parameters

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `n` | integer | `200` | Total number of observations |
| `design` | character | `"mixed"` | One of `"mixed"`, `"continuous"`, `"categorical"` |
| `noise` | numeric | `0.5` | Standard deviation of Gaussian noise |
| `seed` | integer or NULL | `NULL` | Random seed for reproducibility |

## Return Value

A `data.frame` with factor columns and a response column `y`. The data-generating
process includes main effects, two-way interactions, and a weaker three-way interaction.

## Design Types

### Mixed: 2 continuous + 1 categorical

```{r sim-mixed}
dat_mixed = sim_factorial(n = 300, design = "mixed", seed = 42)
str(dat_mixed)
summary(dat_mixed)
```

### Continuous: 3 continuous factors

```{r sim-continuous}
dat_cont = sim_factorial(n = 300, design = "continuous", seed = 42)
str(dat_cont)
```

### Categorical: 3 categorical factors

```{r sim-categorical}
dat_cat = sim_factorial(n = 300, design = "categorical", seed = 42)
str(dat_cat)
```

---

# interaction_surface

The primary function. Generates an interactive 3D surface plot from a fitted model.
Two focal variables map to x and y; the predicted response maps to z. Additional
conditioning factors (`facet_by`) generate separate surfaces — where surfaces cross
indicates interaction effects.

## Parameters

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `model` | model object | *required* | A fitted model with a `predict` method (`lm`, `glm`, `gam`, etc.) |
| `x` | character | *required* | Variable name mapped to the x-axis |
| `y` | character | *required* | Variable name mapped to the y-axis |
| `facet_by` | character or NULL | `NULL` | Variable(s) whose levels generate separate surfaces |
| `n` | integer | `50` | Grid resolution per continuous axis |
| `n_bins` | integer | `3` | Number of bins when `facet_by` is continuous |
| `bin_method` | character | `"quantile"` | Binning method: `"quantile"`, `"equal"`, or `"pretty"` |
| `alpha` | numeric | `0.6` | Surface opacity (0 = transparent, 1 = opaque) |
| `show_points` | logical | `FALSE` | Overlay observed data points, color-coded by `facet_by` level |
| `show_crossings` | logical | `TRUE` | Mark regions where surfaces cross with red markers |
| `show_contour` | logical | `FALSE` | Project crossing points onto the x-y floor plane |
| `contour_z` | numeric or NULL | `NULL` | Custom z-value for contour projection floor (NULL = auto) |
| `labs` | named list or NULL | `NULL` | Axis labels: `list(x = "...", y = "...", z = "...")` |
| `title` | character or NULL | `NULL` | Plot title (auto-generated if NULL) |
| `theme` | character | `"default"` | Color theme: `"default"`, `"viridis"`, or `"grey"` |

## Return Value

A `plotly` htmlwidget with an `ixsurface_meta` attribute containing:

- `x`, `y`, `facet_by` — variable names used
- `n_surfaces` — number of surfaces rendered
- `surface_labels` — character vector of surface names
- `z_matrices` — list of prediction matrices (rows = y, cols = x)
- `x_vals`, `y_vals` — grid coordinate vectors
- `binned_by` — bin info for any continuous `facet_by` variables

## Example 1: Mixed Design — Full Features

```{r surface-mixed, fig.height=7}
fit_mixed = lm(y ~ temp * pressure * catalyst, data = dat_mixed)

interaction_surface(
  fit_mixed,
  x = "temp", y = "pressure", facet_by = "catalyst",
  show_points = TRUE, show_crossings = TRUE, show_contour = TRUE,
  alpha = 0.5,
  labs = list(x = "Temperature (C)", y = "Pressure (psi)", z = "Yield"),
  title = "Mixed Design: temp x pressure | catalyst"
)
```

Three surfaces, one per catalyst level. Red markers show where surfaces cross.
Red circles on the floor are the contour projection. Colored scatter points are
observed data, matched to their nearest surface.

## Example 2: Single Surface (No facet_by)

When `facet_by` is NULL, a single response surface is plotted. Non-focal
variables are held at their median (continuous) or mode (categorical).

```{r surface-single, fig.height=7}
interaction_surface(
  fit_mixed,
  x = "temp", y = "pressure",
  show_points = TRUE, alpha = 0.7,
  title = "Single surface: catalyst held at mode"
)
```

## Example 3: Custom Labels and Low Resolution

```{r surface-lowres, fig.height=7}
interaction_surface(
  fit_mixed,
  x = "temp", y = "pressure", facet_by = "catalyst",
  n = 15, alpha = 0.4,
  labs = list(x = "Temp (C)", y = "Press (psi)", z = "Yield (%)"),
  title = "Low resolution (n = 15)"
)
```

Lower `n` values render faster but produce coarser surfaces.

---

# GLM Support

For `glm` and `gam` models, predictions are automatically returned on the response
scale via `predict(..., type = "response")`. For logistic regression, surfaces
represent probabilities bounded to [0, 1].

```{r surface-glm, fig.height=7}
dat_mixed$success = rbinom(nrow(dat_mixed), 1, plogis((dat_mixed$y - 50) / 4))
gfit = glm(success ~ temp * pressure * catalyst,
           data = dat_mixed, family = binomial)

p_glm = interaction_surface(
  gfit,
  x = "temp", y = "pressure", facet_by = "catalyst",
  show_crossings = TRUE, show_contour = TRUE, alpha = 0.5,
  labs = list(z = "P(success)"),
  title = "GLM: P(success) ~ temp x pressure | catalyst"
)
p_glm
```

### Verify Probability Bounds

```{r glm-bounds}
meta = attr(p_glm, "ixsurface_meta")
z_range = range(unlist(lapply(meta$z_matrices, c)), na.rm = TRUE)
cat("GLM z-range:", round(z_range, 4), "\n")
cat("Bounded to [0, 1]:", z_range[1] >= 0 && z_range[2] <= 1, "\n")
```

---

# find_crossings

Identifies where predicted response surfaces cross for different levels of
conditioning factors. Returns a data frame of approximate crossing locations
without generating a plot.

## Parameters

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `model` | model object | *required* | A fitted model with a `predict` method |
| `x` | character | *required* | First focal variable |
| `y` | character | *required* | Second focal variable |
| `facet_by` | character | *required* | Conditioning variable(s) |
| `n` | integer | `50` | Grid resolution |
| `n_bins` | integer | `3` | Bins for continuous `facet_by` |
| `bin_method` | character | `"quantile"` | Binning method |
| `tolerance` | numeric or NULL | `NULL` | Crossing detection tolerance (NULL = adaptive: 2% of z-range) |

## Return Value

A `data.frame` with columns:

| Column | Type | Description |
|--------|------|-------------|
| `cx` | numeric | x-coordinate of crossing |
| `cy` | numeric | y-coordinate of crossing |
| `cz` | numeric | Predicted response at crossing (average of both surfaces) |
| `pair_label` | character | Which surface pair crosses (e.g., `"catalyst=A vs catalyst=B"`) |

## Example

```{r find-crossings}
cx = find_crossings(fit_mixed, "temp", "pressure", "catalyst")
cat("Total crossing points:", nrow(cx), "\n")
cat("Surface pairs:\n")
print(table(cx$pair_label))
cat("\nFirst rows:\n")
head(cx)
```

```{r find-crossings-summary}
# Summary statistics by pair
for (pair in unique(cx$pair_label)) {
  sub = cx[cx$pair_label == pair, ]
  cat(sprintf("\n%s:\n  x range: [%.1f, %.1f]\n  y range: [%.1f, %.1f]\n  z range: [%.2f, %.2f]\n",
              pair,
              min(sub$cx), max(sub$cx),
              min(sub$cy), max(sub$cy),
              min(sub$cz), max(sub$cz)))
}
```

---

# plot_crossings

Visualizes only the crossing points between interaction surfaces as an interactive
3D scatter plot, color-coded by surface pair. Strips away the surfaces to isolate
where interaction effects are strongest.

## Parameters

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `model` | model object | *required* | A fitted model with a `predict` method |
| `x` | character | *required* | First focal variable |
| `y` | character | *required* | Second focal variable |
| `facet_by` | character | *required* | Conditioning variable(s) |
| `n` | integer | `50` | Grid resolution |
| `n_bins` | integer | `3` | Bins for continuous `facet_by` |
| `bin_method` | character | `"quantile"` | Binning method |
| `tolerance` | numeric or NULL | `NULL` | Crossing detection tolerance |
| `labs` | named list or NULL | `NULL` | Axis labels |
| `title` | character or NULL | `NULL` | Plot title |
| `marker_size` | numeric | `3` | Marker size |
| `marker_opacity` | numeric | `0.7` | Marker opacity |

## Return Value

A `plotly` htmlwidget. If no crossings are found, returns an empty plot with
a "No crossings detected" annotation.

## Example 1: Mixed Design

```{r plot-crossings-mixed, fig.height=7}
plot_crossings(
  fit_mixed, "temp", "pressure", "catalyst",
  labs = list(x = "Temperature (C)", y = "Pressure (psi)", z = "Yield"),
  title = "Mixed: crossing points by surface pair"
)
```

## Example 2: GLM Crossings

```{r plot-crossings-glm, fig.height=7}
plot_crossings(
  gfit, "temp", "pressure", "catalyst",
  labs = list(z = "P(success)"),
  title = "GLM: crossing points (probability scale)",
  marker_size = 4, marker_opacity = 0.8
)
```

## Example 3: Continuous facet_by

```{r plot-crossings-cont, fig.height=7}
fit_cont = lm(y ~ temp * pressure * speed, data = dat_cont)
plot_crossings(
  fit_cont, "temp", "pressure", "speed",
  n_bins = 4, bin_method = "quantile",
  title = "Continuous: crossings by speed-bin pair"
)
```

---

# interaction_surface_grid

Generates interaction surface plots for every pair of focal variables in a model.
Remaining variables become `facet_by` conditioning factors.

## Parameters

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `model` | model object | *required* | A fitted model |
| `factors` | character or NULL | `NULL` | Variables to consider (NULL = all predictors) |
| `facet_max` | integer | `2` | Max number of `facet_by` variables per plot |
| `n` | integer | `30` | Grid resolution (lower default for speed) |
| `n_bins` | integer | `3` | Bins for continuous `facet_by` |
| `alpha` | numeric | `0.6` | Surface opacity |
| `...` | | | Additional arguments passed to `interaction_surface` |

## Return Value

A named list of `plotly` objects. Names use the pattern `"x__y"` (double underscore).
For k factors, produces C(k, 2) plots.

## Example

```{r grid, fig.height=7}
all_plots = interaction_surface_grid(fit_mixed, n = 20, show_points = TRUE)
cat("Number of plots:", length(all_plots), "\n")
cat("Plot names:", paste(names(all_plots), collapse = ", "), "\n")
```

```{r grid-1, fig.height=7}
all_plots$temp__pressure
```

```{r grid-2, fig.height=7}
all_plots$temp__catalyst
```

```{r grid-3, fig.height=7}
all_plots$pressure__catalyst
```

---

# make_prediction_grid

Builds the prediction grid used internally by `interaction_surface` and
`find_crossings`. Exported for advanced use cases where you need direct
access to the grid.

## Parameters

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `model` | model object | *required* | A fitted model |
| `x` | character | *required* | First focal variable |
| `y` | character | *required* | Second focal variable |
| `facet_by` | character or NULL | `NULL` | Conditioning variable(s) |
| `n` | integer | `50` | Grid density for continuous axes |
| `n_bins` | integer | `3` | Bins for continuous `facet_by` |
| `bin_method` | character | `"quantile"` | Binning method |

## Return Value

A list with:

| Component | Type | Description |
|-----------|------|-------------|
| `grid` | data.frame | Prediction-ready data frame with all model variables |
| `binned_by` | named list | Bin info for continuous `facet_by` variables (levels, midpoints, factor column) |

## Behavior

- **Focal variables** (`x`, `y`): regular sequence for continuous; all levels for categorical
- **`facet_by` variables**: all levels for categorical; binned midpoints for continuous
- **Background variables**: held at median (continuous) or mode (categorical)

## Example

```{r pred-grid}
result = make_prediction_grid(fit_mixed, x = "temp", y = "pressure",
                              facet_by = "catalyst", n = 10)
cat("Grid dimensions:", nrow(result$grid), "rows x", ncol(result$grid), "cols\n")
cat("Expected: 10 temp x 10 pressure x 3 catalyst =", 10 * 10 * 3, "\n")
str(result$grid)
```

### With Continuous facet_by (Binned)

```{r pred-grid-binned}
result2 = make_prediction_grid(fit_cont, x = "temp", y = "pressure",
                               facet_by = "speed", n = 10, n_bins = 4)
cat("Grid dimensions:", nrow(result2$grid), "rows x", ncol(result2$grid), "cols\n")
cat("Binned variable:", names(result2$binned_by), "\n")
cat("Bin levels:", result2$binned_by$speed$levels, "\n")
cat("Bin midpoints:", round(result2$binned_by$speed$midpoints, 1), "\n")
```

---

# bin_continuous

Bins a continuous variable into discrete levels. Used internally when a continuous
variable appears in `facet_by`, but exported for standalone use.

## Parameters

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `x` | numeric | *required* | Vector to bin |
| `n_bins` | integer | `3` | Number of bins |
| `method` | character | `"quantile"` | `"quantile"` (equal-count), `"equal"` (equal-width), or `"pretty"` (round breaks) |

## Return Value

A `factor` with descriptive interval labels (from `cut()`).

## Example

```{r bin-continuous}
set.seed(1)
x = rnorm(200, mean = 50, sd = 15)

b_q = bin_continuous(x, n_bins = 3, method = "quantile")
b_e = bin_continuous(x, n_bins = 4, method = "equal")
b_p = bin_continuous(x, n_bins = 3, method = "pretty")

cat("Quantile (3 bins):\n"); print(table(b_q))
cat("\nEqual-width (4 bins):\n"); print(table(b_e))
cat("\nPretty (3 bins):\n"); print(table(b_p))
```

---

# Categorical Axes

When focal variables (`x` or `y`) are categorical or have 6 or fewer unique values,
the surface is constructed over an integer grid with axis tick labels showing level
names rather than numeric indices.

```{r categorical-axes, fig.height=7}
fit_cat = lm(y ~ catalyst * operator * shift, data = dat_cat)

interaction_surface(
  fit_cat,
  x = "catalyst", y = "operator", facet_by = "shift",
  show_points = TRUE, show_crossings = TRUE, alpha = 0.6,
  title = "Fully categorical: axis labels show factor levels"
)
```

```{r categorical-crossings, fig.height=7}
plot_crossings(
  fit_cat, "catalyst", "operator", "shift",
  title = "Categorical crossings: shift=Day vs shift=Night"
)
```

---

# Metadata Access

Every `interaction_surface` plot carries an `ixsurface_meta` attribute for
programmatic downstream use.

```{r metadata}
p = interaction_surface(fit_mixed, x = "temp", y = "pressure",
                        facet_by = "catalyst", n = 20)
meta = attr(p, "ixsurface_meta")

cat("Focal variables:", meta$x, "x", meta$y, "\n")
cat("Facet by:", meta$facet_by, "\n")
cat("Number of surfaces:", meta$n_surfaces, "\n")
cat("Surface labels:", paste(meta$surface_labels, collapse = ", "), "\n")
cat("Z matrix dimensions:", nrow(meta$z_matrices[[1]]), "x", ncol(meta$z_matrices[[1]]), "\n")
cat("X grid range:", range(meta$x_vals), "\n")
cat("Y grid range:", range(meta$y_vals), "\n")
```
