---
title: "ixsurface Visual Exploration"
output: html_document
---

```{r setup, include=FALSE}
install.packages(".", repos = NULL, type = "source", quiet = TRUE)
library(ixsurface)
```

## 1. Mixed Design — Crossing Surfaces with Observed Data

Two continuous factors (temp, pressure) conditioned on a categorical factor (catalyst).
Surfaces cross where interaction effects dominate.

```{r mixed, fig.height=6}
dat = sim_factorial(n = 400, design = "mixed", seed = 7)
fit = lm(y ~ temp * pressure * catalyst, data = dat)

interaction_surface(
  fit, x = "temp", y = "pressure", facet_by = "catalyst",
  show_points = TRUE, show_crossings = TRUE, show_contour = TRUE,
  alpha = 0.5,
  labs = list(x = "Temperature (C)", y = "Pressure (psi)", z = "Yield"),
  title = "Catalyst Interaction: surfaces cross where effects dominate"
)
```

### Crossings Only

```{r mixed-cx, fig.height=6}
plot_crossings(fit, "temp", "pressure", "catalyst",
               labs = list(x = "Temperature (C)", y = "Pressure (psi)", z = "Yield"),
               title = "Mixed: crossing points by surface pair")
```

## 2. GLM — Probability Surfaces Bounded [0, 1]

Logistic regression predictions on the response scale. Surfaces show P(success)
for each catalyst level.

```{r glm, fig.height=6}
dat$success = rbinom(nrow(dat), 1, plogis((dat$y - 50) / 4))
gfit = glm(success ~ temp * pressure * catalyst, data = dat, family = binomial)

interaction_surface(
  gfit, x = "temp", y = "pressure", facet_by = "catalyst",
  show_crossings = TRUE, show_contour = TRUE, alpha = 0.5,
  labs = list(z = "P(success)"),
  title = "Logistic: probability surfaces by catalyst"
)
```

### Crossings Only

```{r glm-cx, fig.height=6}
plot_crossings(gfit, "temp", "pressure", "catalyst",
               labs = list(z = "P(success)"),
               title = "GLM: crossing points by surface pair")
```

## 3. Continuous Facet — Speed Auto-Binned

All three factors are continuous. Speed is automatically binned into 4 quantile
groups to produce separate surfaces.

```{r continuous, fig.height=6}
dat2 = sim_factorial(n = 400, design = "continuous", seed = 7)
fit2 = lm(y ~ temp * pressure * speed, data = dat2)

interaction_surface(
  fit2, x = "temp", y = "pressure", facet_by = "speed",
  n_bins = 4, bin_method = "quantile",
  show_crossings = TRUE, show_contour = TRUE, alpha = 0.5,
  title = "Continuous facet: speed auto-binned into 4 quantile groups"
)
```

### Crossings Only

```{r continuous-cx, fig.height=6}
plot_crossings(fit2, "temp", "pressure", "speed",
               n_bins = 4, bin_method = "quantile",
               title = "Continuous: crossing points by speed-bin pair")
```

## 4. Fully Categorical Design

All factors are discrete. Axis tick labels show level names.

```{r categorical, fig.height=6}
dat3 = sim_factorial(n = 400, design = "categorical", seed = 7)
fit3 = lm(y ~ catalyst * operator * shift, data = dat3)

interaction_surface(
  fit3, x = "catalyst", y = "operator", facet_by = "shift",
  show_points = TRUE, show_crossings = TRUE, alpha = 0.6,
  title = "Categorical: catalyst x operator | shift"
)
```

### Crossings Only

```{r categorical-cx, fig.height=6}
plot_crossings(fit3, "catalyst", "operator", "shift",
               title = "Categorical: crossing points by shift pair")
```

## 5. Single Surface (No Facet)

When `facet_by` is omitted, a single response surface is plotted.
Non-focal variables are held at their median/mode. No crossings possible with
a single surface.

```{r single, fig.height=6}
interaction_surface(
  fit, x = "temp", y = "pressure",
  show_points = TRUE, alpha = 0.7,
  title = "Single surface: temp x pressure (catalyst held at mode)"
)
```

## 6. Pairwise Grid

`interaction_surface_grid` generates all C(k, 2) pairwise plots automatically.

```{r grid, fig.height=6}
grid_plots = interaction_surface_grid(fit, n = 20)
grid_plots$temp__pressure
```

```{r grid2, fig.height=6}
grid_plots$temp__catalyst
```

```{r grid3, fig.height=6}
grid_plots$pressure__catalyst
```
